---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# comma

<!-- badges: start -->
<!-- badges: end -->

**comma** (Comparative Methylomics for Microbial Analysis) is an R package for
analyzing bacterial DNA methylation from Oxford Nanopore sequencing data. It
provides:

- A unified data object (`commaData`) that stores methylation beta values,
  sequencing coverage, sample metadata, and genomic annotations together.
- Parsers for **modkit pileup** BED (primary), Dorado BAM (stub), and Megalodon
  (legacy) output.
- Vectorized annotation of methylation sites against genomic features using
  `GenomicRanges::findOverlaps()`.
- Genome-wide sliding window smoothing for any organism.
- Per-sample methylome quality-control summaries.
- A path toward differential methylation analysis (Phase 4, v0.4.0).

`comma` works with **any sequenced bacterial genome** — no organism-specific
values are hardcoded anywhere.

## Installation

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("carl-stone/CoMMA")
```

## Quick Start

### 1. Construct a `commaData` object

```r
library(comma)

# Named vector: sample_name → path to modkit pileup BED file
files <- c(
  ctrl_1  = "path/to/ctrl_rep1.bed",
  ctrl_2  = "path/to/ctrl_rep2.bed",
  treat_1 = "path/to/treat_rep1.bed"
)

# Sample metadata (data.frame with at least sample_name, condition, replicate)
col_data <- data.frame(
  sample_name = c("ctrl_1",   "ctrl_2",   "treat_1"),
  condition   = c("control",  "control",  "treatment"),
  replicate   = c(1L,         2L,         1L)
)

obj <- commaData(
  files        = files,
  colData      = col_data,
  genome       = "path/to/genome.fa",   # FASTA, BSgenome, or named integer vector
  annotation   = "path/to/genes.gff3",  # GFF3 or BED (optional)
  motif        = "GATC",                # regex motif (optional)
  min_coverage = 5L,
  caller       = "modkit"
)

obj
#> class: commaData
#> sites: 3842 | samples: 3
#> mod types: 6mA, 5mC
#> conditions: control, treatment
#> genome: 1 chromosome (4,641,652 bp total)
#> annotation: 4497 features
#> motif sites: 19120 sites
```

### 2. Access data

```r
# Beta value matrix (sites × samples)
m <- methylation(obj)
dim(m)  # [1] 3842    3

# Coverage matrix
cov <- coverage(obj)

# Per-site metadata (chrom, position, strand, mod_type, ...)
head(siteInfo(obj))

# Per-sample metadata
sampleInfo(obj)

# Modification types present
modTypes(obj)  # "5mC" "6mA"

# Subset to 6mA sites only
obj_6mA <- subset(obj, mod_type = "6mA")
```

### 3. Built-in example data

A synthetic dataset is bundled for testing and examples:

```{r load_example}
library(comma)
data(comma_example_data)
comma_example_data
```

The example object has 300 sites (200 × 6mA, 100 × 5mC), 3 samples
(ctrl\_1, ctrl\_2, treat\_1), and a simulated 100 kb genome.

### 4. Annotate methylation sites

`annotateSites()` uses `GenomicRanges::findOverlaps()` — no nested
for-loops — so it runs in milliseconds even on large genomes.

```{r annotate}
# Overlap annotation: assign feature type and name to each site
annotated <- annotateSites(comma_example_data, type = "overlap")
head(siteInfo(annotated)[, c("chrom", "position", "mod_type",
                              "feature_type", "feature_name")])
```

```{r metagene}
# Metagene annotation: fractional position within feature (0=start, 1=end)
mg <- annotateSites(comma_example_data, type = "metagene")
head(siteInfo(mg)[, c("position", "metagene_feature", "metagene_pos")])
```

```{r proximity}
# Proximity annotation: distance and signed offset from nearest TSS
prox <- annotateSites(comma_example_data, type = "proximity", window = 5000L)
head(siteInfo(prox)[, c("position", "nearest_feature",
                         "distance_to_feature", "rel_pos")])
```

### 5. Sliding window methylation

`slidingWindow()` smooths methylation along the genome for each sample. Genome
size comes from `genome(object)` — never hardcoded.

```{r sliding_window}
sw <- slidingWindow(comma_example_data, window = 5000L, stat = "median")
head(sw)
```

Visualize with ggplot2:

```{r sliding_window_plot, eval=FALSE}
library(ggplot2)
ggplot(sw[sw$sample_name == "ctrl_1", ],
       aes(x = position, y = window_median)) +
  geom_line() +
  labs(x = "Genome position", y = "Median methylation (5 kb window)",
       title = "ctrl_1 — genome-wide methylation profile") +
  theme_bw()
```

### 6. Methylome summary

`methylomeSummary()` produces per-sample QC statistics:

```{r methylome_summary}
ms <- methylomeSummary(comma_example_data)
ms[, c("sample_name", "condition", "n_covered", "mean_beta", "frac_methylated")]
```

### 7. Coverage analysis

```{r coverage_depth}
# Windowed depth across the genome
cd <- coverageDepth(comma_example_data, window = 10000L)
head(cd)
```

```{r variance_by_depth}
# Methylation variance as a function of sequencing depth
vd <- varianceByDepth(comma_example_data, coverage_bins = 5:30)
head(vd[!is.na(vd$variance), ])
```

### 8. Export to BED

```{r writeBED, eval=FALSE}
writeBED(comma_example_data,
         file     = "ctrl_1_6mA.bed",
         sample   = "ctrl_1",
         mod_type = "6mA")
```

## Roadmap

| Version | Phase | Status |
|---|---|---|
| 0.2.0 | Data infrastructure — `commaData` S4 class, modkit parser | ✅ Complete |
| 0.3.0 | Refactored functions — `annotateSites`, `slidingWindow`, `methylomeSummary` | ✅ Complete |
| 0.4.0 | Differential methylation — `diffMethyl()`, beta-binomial model | ⏳ Next |
| 0.5.0 | Visualization & Bioconductor preparation | ⏳ Pending |
| 1.0.0 | Bioconductor submission | ⏳ Pending |

## Citation

Stone CJ et al. (2022). Genome-wide adenine methylation in *Escherichia coli*
K-12 reveals methylation patterns associated with gene regulation. *bioRxiv*.
