---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	fig.path = "man/figures/README-",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	out.width = "100%"
)
```


# comma

<!-- badges: start -->
[![R-CMD-check](https://github.com/carl-stone/CoMMA/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/carl-stone/CoMMA/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**comma** (**co**mma: for **m**icrobial **m**ethylation **a**nalysis) is
an R package for genome-wide analysis of bacterial DNA methylation from Oxford
Nanopore sequencing data. It accepts output from **modkit**, **Dorado**, and
**Megalodon**, supports **6mA, 5mC, and 4mC** in a single unified container,
and covers the full analysis workflow from raw files through quality control,
site annotation, differential methylation testing, and publication-quality
visualization.

## Features

- **`commaData` S4 class** — extends `SummarizedExperiment`; stores beta
  values, coverage, sample metadata, genome info, and annotations together.
- **Three parsers** — modkit pileup BED (primary), Dorado BAM (full MM/ML tag
  decoding), and Megalodon (legacy).
- **Vectorized annotation** via `GenomicRanges::findOverlaps()`; three modes:
  overlap, proximity, and metagene (normalized position within feature bodies).
- **Differential methylation** — quasibinomial GLM per site with a
  DESeq2-style interface (`diffMethyl()` → `results()` → `filterResults()`).
- **Seven `plot_*()` functions** — coverage QC, beta distributions, PCA, genome
  tracks, metagene profiles, volcano plots, and heatmaps; all return `ggplot`
  objects for further customization.
- **Any bacterial genome** — no organism-specific values are hardcoded anywhere.
- **Multi-modification-type** — 6mA, 5mC, and 4mC coexist in one object;
  every function accepts a `mod_type` argument to work with individual types.

## Installation

```{r}
# Development version from GitHub:
devtools::install_github("carl-stone/CoMMA")

# Bioconductor (upcoming v1.0.0):
# BiocManager::install("comma")
```

## The `commaData` Object

`commaData` extends `SummarizedExperiment` and is constructed from per-sample
methylation files:

```{r eval=FALSE}
library(comma)

# Named vector: sample_name → path to modkit pileup BED
files <- c(
  ctrl_1  = "path/to/ctrl_rep1.bed",
  ctrl_2  = "path/to/ctrl_rep2.bed",
  treat_1 = "path/to/treat_rep1.bed",
  treat_2 = "path/to/treat_rep2.bed"
)

col_data <- data.frame(
  sample_name = c("ctrl_1",    "ctrl_2",    "treat_1",    "treat_2"),
  condition   = c("control",   "control",   "treatment", "treatment"),
  replicate   = c(1L,          2L,          1L,          2L)
)

obj <- commaData(
  files        = files,
  colData      = col_data,
  genome       = "path/to/genome.fa",   # FASTA, BSgenome, or named integer vector
  annotation   = "path/to/genes.gff3",  # GFF3 or BED (optional)
  motif        = "GATC",                # regex motif (optional)
  min_coverage = 5L,
  caller       = "modkit"               # "modkit", "dorado", or "megalodon"
)
```

A built-in synthetic dataset (300 sites, 3 samples, 100 kb genome) is included
for testing and demonstrations:

```{r load-data}
library(comma)
data(comma_example_data)
comma_example_data
```

## Workflow

### Step 1 — Quality Control

`methylomeSummary()` returns per-sample distribution statistics:

```{r qc-summary}
ms <- methylomeSummary(comma_example_data)
ms[, c("sample_name", "condition", "mean_beta", "median_beta",
       "frac_methylated", "n_covered")]
```

`plot_coverage()` shows the sequencing depth distribution per sample:

```{r plot-coverage, fig.alt = "Coverage depth distribution per sample."}
plot_coverage(comma_example_data)
```

`plot_methylation_distribution()` plots the density of beta values, faceted by
modification type. Bacterial methylomes often show a bimodal distribution
(sites are either fully methylated or unmethylated):

```{r plot-dist, fig.alt = "Beta value density per sample, faceted by modification type."}
plot_methylation_distribution(comma_example_data)
```

`plot_pca()` runs PCA on per-sample methylation profiles for sample-level QC:

```{r plot-pca, fig.alt = "PCA of methylation profiles colored by condition."}
plot_pca(comma_example_data, color_by = "condition")
```

### Step 2 — Annotate Sites

`annotateSites()` maps sites to genomic features using
`GenomicRanges::findOverlaps()` — vectorized, no nested loops:

```{r annotate}
annotated <- annotateSites(comma_example_data, type = "overlap")

# Fraction of sites overlapping at least one annotated feature
si <- siteInfo(annotated)
mean(lengths(si$feature_names) > 0)
```

Three annotation modes are available: `"overlap"`, `"proximity"` (nearest
feature with signed distance), and `"metagene"` (normalized position 0–1
within feature bodies).

`plot_metagene()` shows the average methylation profile across gene bodies:

```{r plot-metagene, fig.alt = "Mean methylation across gene bodies (TSS = 0 to TTS = 1)."}
plot_metagene(comma_example_data, feature = "gene")
```

### Step 3 — Genome Track Visualization

`plot_genome_track()` produces a genome browser–style scatter plot of
methylation along a chromosomal region:

```{r plot-track, fig.alt = "Genome track for the first 50 kb of chr_sim, 6mA sites."}
plot_genome_track(comma_example_data, chromosome = "chr_sim",
                  start = 1L, end = 50000L, mod_type = "6mA")
```

### Step 4 — Differential Methylation

`diffMethyl()` tests each site for differential methylation between conditions.
It is modeled on DESeq2's workflow: pass a `commaData` object and a design
formula, get back the same object with per-site statistics in `rowData`:

```{r diff-methyl}
cd_dm <- diffMethyl(comma_example_data, formula = ~ condition,
                    mod_type = "6mA")
```

Extract results as a tidy data frame and filter to significant sites:

```{r results}
res <- results(cd_dm)
sig <- filterResults(cd_dm, padj = 0.05, delta_beta = 0.2)
cat("Total 6mA sites tested:", nrow(res), "\n")
cat("Significant sites (padj < 0.05, |Δβ| ≥ 0.2):", nrow(sig), "\n")

# Top hits
head(res[order(res$dm_padj),
         c("chrom", "position", "dm_delta_beta", "dm_padj")])
```

`plot_volcano()` displays the differential methylation landscape — effect size
(delta methylation) versus significance (–log10 padj):

```{r plot-volcano, fig.alt = "Volcano plot of differential 6mA methylation."}
plot_volcano(res)
```

`plot_heatmap()` shows per-sample beta values for the top differentially
methylated sites:

```{r plot-heatmap, fig.alt = "Heatmap of top 30 differentially methylated 6mA sites."}
plot_heatmap(res, cd_dm, n_sites = 30L)
```

### Step 5 — Export to BED

```{r writeBED, eval = FALSE}
writeBED(cd_dm, file = "results_6mA.bed", sample = "ctrl_1", mod_type = "6mA")
```

## Multi-Modification-Type Support

A single `commaData` object can hold 6mA, 5mC, and 4mC sites simultaneously.
The `mod_type` argument is available on every analysis function:

```{r mod-types}
# Modification types present in the example data
modTypes(comma_example_data)

# Subset to a single type
obj_6mA <- subset(comma_example_data, mod_type = "6mA")
obj_6mA
```

For a complete joint 6mA + 5mC analysis, see the
**Multiple Modification Types** vignette (`vignette("multiple-modification-types", package = "comma")`).

## Documentation

```{r eval=FALSE, include=TRUE}
?comma        # Package overview and five-step workflow
?commaData    # Constructor and commaData class
?diffMethyl   # Differential methylation testing
```

Two vignettes are included:

- **Getting Started** (`vignette("getting-started", package = "comma")`) —
  end-to-end workflow: load → QC → annotate → differential methylation →
  visualize.
- **Multiple Modification Types**
  (`vignette("multiple-modification-types", package = "comma")`) — joint 6mA
  and 5mC analysis in a single `commaData` object.

## Roadmap

| Version | Phase | Status |
|---|---|---|
| 0.2.0 | Data infrastructure — `commaData` S4 class, modkit parser | ✅ Complete |
| 0.3.0 | Refactored analysis — `annotateSites`, `slidingWindow`, QC utilities | ✅ Complete |
| 0.4.0 | Differential methylation — `diffMethyl()`, beta-binomial GLM | ✅ Complete |
| 0.5.0 | Visualization — seven `plot_*()` functions, vignettes, package docs | ✅ Complete |
| 1.0.0 | Bioconductor submission | ⏳ In preparation |

## Citation

Stone CJ et al. (2022). Genome-wide adenine methylation in *Escherichia coli*
K-12 reveals methylation patterns associated with gene regulation. *bioRxiv*.
