---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CoMMA

<!-- badges: start -->
<!-- badges: end -->

CoMMA is an R package for comparative bacterial methylomics. It standardizes per-site methylation calls, annotates sites with genomic features, and runs differential methylation analyses with manuscript-aligned defaults.

## Installation

You can install the development version of CoMMA from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("carl-stone/CoMMA")
```

## Quickstart (runtime-friendly)

```{r setup}
library(CoMMA)
```

### 1) Minimal ingest example

Use one of CoMMA's pragmatic import helpers to map caller outputs into the canonical site schema, then validate with `validate_site_table()`.

```{r ingest-example}
modkit_df <- data.frame(
  chrom = c("chr1", "chr1"),
  start = c(99L, 119L),
  strand = c("+", "+"),
  modified_primary_base = c("6mA", "6mA"),
  n_mod = c(16L, 4L),
  valid_coverage = c(20L, 18L)
)

validated_sites <- convert_modkit_bedmethyl(
  modkit_df,
  sample_id = "S1",
  group = "WT"
)
validated_sites
```


### Supported ingest templates

CoMMA currently ships pragmatic converters for these common input templates:

- `convert_modkit_bedmethyl()`
  - expected columns: `chrom`/`seqname`, `start`/`pos`, `strand`,
    `modified_primary_base`/`mod_base`, `n_mod`, `valid_coverage`/`n_total`
  - assumes `start` is 0-based if provided and converts to 1-based `pos`
- `convert_megalodon_tsv()`
  - expected columns: sequence `chromosome`/`contig`/`seqname`,
    position `position`/`pos`, `strand`, modified base `mod_base`/`modified_base`,
    modified count `n_mod`/`mod_count`, total count `n_total`/`coverage`/`read_depth`
  - supports optional `motif`; use `position_is_zero_based = TRUE` for 0-based tables
- `convert_beta_coverage_long()`
  - expected columns: `seqname`, `pos`, `strand`, `beta`, `coverage`,
    `sample_id`, `group`
  - optional columns: `mod_base`, `motif`
  - supports `position_is_zero_based = TRUE` and non-6mA contexts (e.g., `5mC`, `4mC`)

All converters end by calling `validate_site_table()`.

### 2) Annotation example (GFF + BED readers)

Read annotation features from either GFF or BED, then map methylation sites to overlapping features with `annotate_sites_with_features()`.

```{r annotation-example}
gff_file <- tempfile(fileext = ".gff")
writeLines(
  c(
    "##gff-version 3",
    "chr1\ttoy\tgene\t90\t140\t.\t+\t.\tID=geneA;Name=geneA"
  ),
  gff_file
)

gff_features <- read_annotation_gff(gff_file)
annotated_gff <- annotate_sites_with_features(validated_sites, gff_features)
annotated_gff[, c("seqname", "pos", "sample_id", "feature_id", "feature_type")]
```

```{r annotation-bed-example}
bed_file <- tempfile(fileext = ".bed")
writeLines(
  c(
    "chr1\t80\t110\tpromoterA",
    "chr1\t115\t130\tpromoterB"
  ),
  bed_file
)

bed_features <- read_annotation_bed(bed_file, feature_type = "promoter")
annotate_sites_with_features(validated_sites, bed_features)[,
  c("seqname", "pos", "sample_id", "feature_id", "feature_type")
]
```

### 3) Differential methylation example

`run_differential_methylation()` uses manuscript defaults by default:

- coverage filter: `n_total >= 10`
- coverage normalization: median
- significance thresholds: `qvalue < 0.05` and `abs(percent_diff) > 10`

The example below includes a non-6mA context (`5mC` at `CCWGG`).

```{r differential-example}
site_table_5mc <- data.frame(
  seqname = rep("chr1", 8),
  pos = rep(c(100L, 120L), each = 4),
  strand = "+",
  mod_base = "5mC",
  motif = "CCWGG",
  n_mod = c(18L, 4L, 19L, 5L, 3L, 17L, 2L, 16L),
  n_total = c(20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L),
  sample_id = c("WT1", "WT1", "WT2", "WT2", "KO1", "KO1", "KO2", "KO2"),
  group = c("WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO")
)

if (requireNamespace("methylKit", quietly = TRUE)) {
  diff_res <- run_differential_methylation(
    site_table_5mc,
    mod_base = "5mC",
    motif = "CCWGG"
  )

  diff_res$result_table
} else {
  message("Install 'methylKit' to run this example.")
}
```

## Data contract

CoMMA's canonical site-table schema requires:

- `seqname`, `pos`, `strand`
- `sample_id`, `group`
- one measurement mode:
  - counts: `n_mod` + `n_total`, or
  - fraction: `beta` + `coverage` (counts are backfilled)
- optional metadata: `mod_base` (defaults to `"6mA"`), `motif` (defaults to `"GATC"`)

`validate_site_table()` returns a standardized table with stable columns:

- `seqname`, `pos`, `strand`, `mod_base`, `motif`
- `n_mod`, `n_total`, `sample_id`, `group`, `beta`

`run_differential_methylation()` returns a list with stable components:

- `result_table` (`seqname`, `pos`, `strand`, `pvalue`, `qvalue`, `percent_diff`, `significant`)
- `plot_data` (`volcano_data`, `beta_by_sample`)
- `methylkit_objects` (`methyl_raw`, `normalized`, `united`, `diff`)


## Repository layout notes

For contributors, package files should live in canonical directories:

- active package code: `R/`, `tests/testthat/`, `vignettes/`, `inst/`
- reproducible example fixtures: `inst/extdata/`
- test-only fixtures: `tests/testthat/testdata/`
- historical non-package analyses: `misc/legacy/`

See `CONTRIBUTING.md` for a short maintainer-facing convention guide.
