---
title: "CoMMA end-to-end (toy example)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CoMMA end-to-end (toy example)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(42)
```

This vignette demonstrates a fast, deterministic workflow using a tiny synthetic dataset:

1. Build CoMMA's canonical site table.
2. Import toy GFF/BED annotations.
3. Annotate methylation sites by overlap.
4. Run differential methylation with manuscript-default thresholds.
5. Produce compact summaries and one plot-ready output.

```{r setup}
library(CoMMA)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) {
  library(ggplot2)
}
```

## 1) Build a canonical site table

```{r toy-site-table}
toy_sites <- data.frame(
  seqname = rep("chr1", 24),
  pos = rep(c(100L, 200L, 300L, 400L, 500L, 600L), times = 4),
  strand = rep(c("+", "-", "+", "-", "+", "-"), times = 4),
  mod_base = "6mA",
  motif = "GATC",
  n_mod = c(
    # WT1
    6L, 9L, 5L, 8L, 6L, 9L,
    # WT2
    7L, 8L, 6L, 9L, 6L, 8L,
    # mut1
    2L, 4L, 2L, 8L, 2L, 4L,
    # mut2
    3L, 3L, 3L, 9L, 3L, 3L
  ),
  n_total = c(
    12L, 12L, 10L, 12L, 12L, 12L,
    12L, 12L, 10L, 12L, 12L, 12L,
    12L, 12L, 10L, 12L, 12L, 12L,
    12L, 12L, 10L, 12L, 12L, 12L
  ),
  sample_id = rep(c("WT1", "WT2", "MMR1", "MMR2"), each = 6),
  group = rep(c("WT", "WT", "MMR", "MMR"), each = 6),
  stringsAsFactors = FALSE
)

site_table <- validate_site_table(toy_sites)
head(site_table)
```

## 2) Import toy GFF and BED annotations

```{r toy-annotations}
gff_file <- tempfile(fileext = ".gff")
writeLines(
  c(
    "chr1\ttoy\tgene\t80\t250\t.\t+\t.\tID=geneA;Name=geneA",
    "chr1\ttoy\tgene\t350\t620\t.\t-\t.\tID=geneB;Name=geneB"
  ),
  con = gff_file
)

gff_features <- read_annotation_gff(gff_file)

bed_file <- tempfile(fileext = ".bed")
writeLines(
  c(
    "chr1\t90\t130\tpromoterA\t0\t+",
    "chr1\t540\t610\tisland1\t0\t-"
  ),
  con = bed_file
)

bed_features <- read_annotation_bed(bed_file, feature_type = "regulatory_region")

feature_table <- rbind(
  gff_features[, c("feature_type", "feature_id", "seqname", "start", "end", "strand")],
  bed_features[, c("feature_type", "feature_id", "seqname", "start", "end", "strand")]
)

feature_table
```

## 3) Annotate sites by overlap

```{r annotate-sites}
annotated_sites <- annotate_sites_with_features(
  site_table = site_table,
  feature_table = feature_table,
  site_pos_col = "pos",
  site_seqname_col = "seqname",
  include_unannotated = TRUE
)

annotation_summary <- as.data.frame(table(annotated_sites$feature_type, useNA = "ifany"))
colnames(annotation_summary) <- c("feature_type", "n_rows")
annotation_summary
```

## 4) Differential methylation with manuscript-default thresholds

Defaults used here match the package contract:

- minimum coverage: `10`
- q-value threshold: `0.05`
- absolute percent difference threshold: `10`
- coverage normalization by median
- SLIM correction in methylKit

```{r run-diff}
diff_out <- run_differential_methylation(
  site_table = site_table,
  mod_base = "6mA",
  motif = "GATC"
)

result_table <- diff_out$result_table
head(result_table)
```

## 5) Compact summaries and one plot-ready output

```{r summaries}
signif_summary <- data.frame(
  n_sites = nrow(result_table),
  n_significant = sum(result_table$significant, na.rm = TRUE),
  median_abs_percent_diff = median(abs(result_table$percent_diff), na.rm = TRUE)
)

signif_summary
```

`run_differential_methylation()` returns plotting-ready volcano data.

```{r volcano-plot, fig.width=5, fig.height=3.5}
volcano_df <- diff_out$plot_data$volcano_data

head(volcano_df[, c("seqname", "pos", "qvalue", "percent_diff", "significant")])

if (has_ggplot2) {
  # Minimal plotting example (can be reused in downstream analyses)
  ggplot(volcano_df, aes(x = percent_diff, y = neg_log10_qvalue, color = significant)) +
    geom_point(size = 2) +
    labs(
      title = "Toy differential methylation volcano",
      x = "Percent methylation difference",
      y = "-log10(q-value)",
      color = "Significant"
    ) +
    theme_minimal(base_size = 10)
} else {
  message("Install ggplot2 to render the volcano plot example.")
}
```
