---
title: "Getting Started with comma"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting Started with comma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    fig.width  = 7,
    fig.height = 5,
    warning    = FALSE,
    message    = FALSE
)
```

# Introduction

`comma` (**Co**mparative **M**ethylomics for **M**icrobial **A**nalysis) is an
R package for genome-wide analysis of bacterial DNA methylation from Oxford
Nanopore sequencing data. It supports three modification types — N6-methyladenine
(6mA), 5-methylcytosine (5mC), and N4-methylcytosine (4mC) — in a single,
unified data container. This vignette walks through the complete analysis
workflow using the built-in `comma_example_data` synthetic dataset.

The typical `comma` workflow has five steps:

1. **Load** per-sample methylation files into a `commaData` object.
2. **QC** the data (coverage, beta distributions, PCA).
3. **Annotate** sites relative to genomic features.
4. **Test** for differential methylation between conditions.
5. **Visualize** the results.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("comma")
```

```{r load}
library(comma)
```

# The `commaData` Object

`commaData` extends `SummarizedExperiment` and is the central data container
in `comma`. It stores:

- **methylation** — a sites × samples matrix of beta values (0–1).
- **coverage** — a sites × samples matrix of read depths.
- **rowData** — per-site metadata: chromosome, position, strand, mod_type.
- **colData** — per-sample metadata: sample_name, condition, replicate.
- **genomeInfo** — chromosome names and lengths.
- **annotation** — genomic features as a `GRanges` object.
- **motifSites** — motif instances as a `GRanges` object.

The built-in `comma_example_data` contains 300 synthetic methylation sites
(200 × 6mA, 100 × 5mC) on a simulated 100 kb chromosome across three samples:
two controls (`ctrl_1`, `ctrl_2`) and one treatment (`treat_1`).

```{r load-data}
data(comma_example_data)
comma_example_data
```

```{r explore-basics}
# Modification types present
modTypes(comma_example_data)

# Per-sample metadata
sampleInfo(comma_example_data)

# Matrix dimensions: sites × samples
dim(methylation(comma_example_data))
```

# Exploring the Methylome

## Summary Statistics

`methylomeSummary()` returns a tidy data frame with per-sample distribution
statistics:

```{r methylome-summary}
ms <- methylomeSummary(comma_example_data)
ms[, c("sample_name", "condition", "mean_beta", "median_beta", "n_covered")]
```

## Coverage QC

`plot_coverage()` shows the distribution of sequencing depth per site, per
sample. Consistent coverage across samples is an important quality indicator.

```{r plot-coverage, fig.cap = "Coverage depth distribution per sample."}
plot_coverage(comma_example_data)
```

## Beta Value Distributions

`plot_methylation_distribution()` plots the density of methylation levels for
each sample. Bacterial genomes often show a bimodal distribution (sites are
either fully methylated or unmethylated).

```{r plot-dist, fig.cap = "Methylation beta value density per sample, faceted by modification type."}
plot_methylation_distribution(comma_example_data)
```

Restrict to a single modification type:

```{r plot-dist-6ma, fig.cap = "Beta value density for 6mA sites only."}
plot_methylation_distribution(comma_example_data, mod_type = "6mA")
```

## PCA for Sample-Level QC

`plot_pca()` performs PCA on per-sample methylation profiles. Samples from
the same condition should cluster together. Internally, beta values are
converted to M-values via `mValues()` before PCA, which stabilizes variance
across sites near 0 or 1.

```{r plot-pca, fig.cap = "PCA of methylation profiles colored by condition."}
plot_pca(comma_example_data, color_by = "condition")
```

To retrieve the underlying scores for custom plotting, use `return_data = TRUE`.
The result is a `data.frame` with `PC1`, `PC2`, and all sample metadata columns;
the percentage of variance explained by each PC is stored in
`attr(result, "percentVar")`.

```{r pca-data}
pca_df <- plot_pca(comma_example_data, return_data = TRUE)
attr(pca_df, "percentVar")  # variance explained by PC1 and PC2
```

# Annotating Sites

`annotateSites()` maps methylation sites to genomic features, always computing
four parallel list columns in `rowData`:

- `feature_types` — GFF3 feature type for each association (e.g., `"gene"`, `"promoter"`).
- `feature_names` — feature name for each association.
- `rel_position` — signed distance from the feature (0 = inside; negative = upstream; positive = downstream, strand-aware).
- `frac_position` — normalized position within the feature ([0, 1]; `NA` for sites outside).

Use the `keep` argument to filter which associations are retained: `"all"`
(default), `"overlap"` (only overlapping features), `"proximity"` (retains
`rel_position`, drops `frac_position`), or `"metagene"` (only overlapping
features with `frac_position`).

```{r annotate}
annotated <- annotateSites(comma_example_data)
si <- siteInfo(annotated)

# Proportion of sites overlapping at least one annotated feature
mean(lengths(si$feature_names) > 0)
```

`plot_metagene()` visualizes the average methylation profile across gene bodies:

```{r plot-metagene, fig.cap = "Mean methylation profile across gene bodies (TSS to TTS)."}
plot_metagene(comma_example_data, feature = "gene")
```

# Genome Track Visualization

`plot_genome_track()` produces a genome browser–style plot of methylation
along a chromosome region:

```{r plot-track, fig.cap = "Genome track for the first 50 kb of chr_sim."}
plot_genome_track(comma_example_data, chromosome = "chr_sim",
                  start = 1L, end = 50000L, mod_type = "6mA")
```

# Differential Methylation

`diffMethyl()` tests each site for differential methylation between conditions.
It is modeled on DESeq2's workflow: pass a `commaData` object and a design
formula, and receive back the same object with statistical results in `rowData`.

```{r diff-methyl}
cd_dm <- diffMethyl(comma_example_data, formula = ~ condition,
                    mod_type = "6mA")
cd_dm
```

Extract the results as a tidy data frame:

```{r results}
res <- results(cd_dm)
# Top sites by adjusted p-value
head(res[order(res$dm_padj),
         c("chrom", "position", "mod_type", "dm_delta_beta", "dm_padj")])
```

Filter to significant sites (padj < 0.05, |Δβ| ≥ 0.2):

```{r filter-results}
sig <- filterResults(cd_dm, padj = 0.05, delta_beta = 0.2)
cat("Significant sites:", nrow(sig), "\n")
```

## Volcano Plot

`plot_volcano()` displays the differential methylation landscape. Sites are
colored by direction and significance:

```{r plot-volcano, fig.cap = "Volcano plot: effect size (Δβ) vs. significance (–log10 padj)."}
plot_volcano(res)
```

## Heatmap of Top Sites

`plot_heatmap()` shows methylation beta values for the top differentially
methylated sites:

```{r plot-heatmap, fig.cap = "Heatmap of top 30 differentially methylated 6mA sites."}
plot_heatmap(res, cd_dm, n_sites = 30L)
```

# Session Information

```{r session-info}
sessionInfo()
```
