---
title: "Troubleshooting Data Import"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Troubleshooting Data Import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)
```

# Overview

Most import problems come from one of four places:

1. The file format does not match the selected `caller`.
2. Sample names in `files` and `colData$sample_name` do not match exactly.
3. Genome information is missing, unnamed, or uses chromosome names that do not
   match the methylation files.
4. Annotation files are missing optional dependencies or use an unexpected file
   extension.

Start by checking that each input sample can be parsed with the same caller and
that the sample names are identical across `files` and `colData`.

```{r minimal-constructor, eval = FALSE}
library(comma)

files <- c(
    ctrl_1 = "ctrl_1.modkit.bed",
    treat_1 = "treat_1.modkit.bed"
)

col_data <- data.frame(
    sample_name = c("ctrl_1", "treat_1"),
    condition = c("control", "treatment"),
    replicate = c(1L, 1L)
)

obj <- commaData(
    files = files,
    colData = col_data,
    genome = c(chr1 = 4641652L),
    caller = "modkit"
)
```

# Choosing The Caller

Use `caller = "modkit"` for modkit pileup BED files. This is the recommended
path for most Dorado-based workflows: call bases with Dorado, run
`modkit pileup`, then import the resulting BED files into `comma`.

Use `caller = "dorado"` only when you want `comma` to parse an aligned BAM file
with MM/ML base modification tags directly. The BAM must contain modification
tags and should be indexed. Direct BAM parsing does not provide motif context,
so per-site `motif` values will be `NA`.

Use `caller = "megalodon"` for legacy Megalodon per-read modification BED
files. Megalodon files do not encode the modification type in a way `comma`
can infer, so provide exactly one `mod_type` value.

```{r caller-examples, eval = FALSE}
# Recommended path for Dorado runs after modkit pileup
obj_modkit <- commaData(files, col_data, genome = c(chr1 = 4641652L),
                        caller = "modkit")

# Direct Dorado BAM import
obj_dorado <- commaData(bam_files, col_data, genome = c(chr1 = 4641652L),
                        caller = "dorado")

# Legacy Megalodon import requires a single explicit modification type
obj_megalodon <- commaData(megalodon_files, col_data,
                           genome = c(chr1 = 4641652L),
                           caller = "megalodon",
                           mod_type = "5mC")
```

# Modkit BED Format Problems

`comma` expects modkit pileup bedMethyl output with at least 18 columns and no
header row. The parser uses the first 18 fields:

- columns 1 to 3: chromosome, 0-based start, and end
- column 4: modification code, often `code,motif,position`
- column 6: strand
- column 10: valid coverage
- column 11: percent modified

The parser converts 0-based BED starts to 1-based genomic positions and converts
percent modified values from 0 to 100 into beta values from 0 to 1.

Common symptoms and fixes:

| Symptom | Likely cause | Fix |
|---|---|---|
| `expected at least 18` | The file is not modkit pileup bedMethyl output, or it was truncated. | Re-run `modkit pileup` and pass the BED output, not the BAM. |
| Unknown `mod_code` warning | Column 4 contains codes outside the supported modkit map. | Confirm the file is a methylation pileup and decide whether the code should map to `6mA`, `5mC`, or `4mC`. |
| No sites remain | Coverage or `mod_type` filtering removed every row. | Lower `min_coverage`, check `mod_type`, and inspect the input file for nonzero coverage. |
| Duplicate methylation site rows | The file has repeated rows for the same chromosome, position, strand, modification type, and motif. | Aggregate or regenerate the per-sample pileup before importing. |

If the modkit `mod_code` field includes motif context, such as `a,GATC,1`,
`comma` stores `mod_type = "6mA"` and `motif = "GATC"`. Older or derived files
without motif context can still import, but `motif` will be `NA`.

```{r inspect-modkit, eval = FALSE}
first_rows <- read.table("ctrl_1.modkit.bed", nrows = 5, fill = TRUE)
ncol(first_rows)
first_rows[, c(1, 2, 3, 4, 6, 10, 11)]
```

# Sample Metadata Problems

`files` must be a named character vector, and its names must match
`colData$sample_name`. The constructor also requires `sample_name`, `condition`,
and `replicate` columns in `colData`.

```{r sample-name-checks, eval = FALSE}
setdiff(names(files), col_data$sample_name)
setdiff(col_data$sample_name, names(files))
```

If either result is non-empty, fix the names before calling `commaData()`.
Additional sample metadata columns are allowed and are preserved in
`sampleInfo(object)`.

# Genome Size Problems

Genome information is strongly recommended because it becomes the `Seqinfo`
attached to methylation sites. Use one of these forms:

- a named integer or numeric vector of chromosome sizes
- a FASTA file path
- a named `Biostrings::DNAStringSet`
- a whole `BSgenome` object

```{r genome-examples, eval = FALSE}
# Named vector: simplest and robust for one bacterial chromosome
genome <- c(NC_000913 = 4641652L)

# FASTA path: names come from FASTA sequence headers
genome <- "E_coli_K12.fa"

# Whole BSgenome object: pass the object, not one chromosome extracted with $
genome <- BSgenome.Ecoli.NCBI.20080805::BSgenome.Ecoli.NCBI.20080805
```

Do not pass a single unnamed sequence such as `BSgenomeObject$NC_000913`.
That creates a `DNAString` without a chromosome name, so `comma` cannot attach
the sequence length to methylation sites. Use the whole `BSgenome` object or a
named vector instead.

Chromosome names must match the methylation files. If the BED file uses `chr1`
but the genome vector is named `NC_000913`, rename one side before import. Extra
chromosomes in the genome input are dropped with a message when they are not
present in the methylation data.

```{r genome-name-checks, eval = FALSE}
bed_chroms <- unique(read.table("ctrl_1.modkit.bed", nrows = 1000)[[1]])
names(genome)
setdiff(bed_chroms, names(genome))
```

# Annotation File Problems

Annotation is optional. Use `annotation = NULL` while debugging methylation
file import, then add annotation once the `commaData` object can be constructed.

`comma` accepts annotation as a `GRanges` object or as a GFF, GFF3, or BED file
path. Loading annotation files requires the Bioconductor package
`rtracklayer`.

```{r annotation-examples, eval = FALSE}
if (requireNamespace("rtracklayer", quietly = TRUE)) {
    genes <- loadAnnotation("genes.gff3", feature_types = "gene")
    obj <- commaData(files, col_data, genome = c(chr1 = 4641652L),
                     annotation = genes, caller = "modkit")
}
```

Common symptoms and fixes:

| Symptom | Likely cause | Fix |
|---|---|---|
| `Package 'rtracklayer' is required` | The optional annotation dependency is not installed. | Install with `BiocManager::install("rtracklayer")`, or set `annotation = NULL`. |
| `Annotation file not found` | The path is wrong relative to the R working directory. | Use `normalizePath()` or an absolute path to confirm the file. |
| Unsupported extension | The file extension is not `.gff`, `.gff3`, or `.bed` after removing compression suffixes. | Rename or convert the file, or import it yourself as `GRanges`. |
| No requested feature types | `feature_types` does not match the standardized feature type column. | Inspect `table(mcols(loadAnnotation(file))$feature_type)`. |

# Quick Checklist

Before opening an issue or debugging further, collect these facts:

```{r checklist, eval = FALSE}
file.exists(files)
names(files)
col_data$sample_name

first_rows <- read.table(files[[1]], nrows = 5, fill = TRUE)
dim(first_rows)
first_rows[, seq_len(min(18, ncol(first_rows)))]

genome
```

Then try a minimal constructor call without annotation or motif scanning:

```{r minimal-debug, eval = FALSE}
obj <- commaData(
    files = files,
    colData = col_data,
    genome = c(chr1 = 4641652L),
    annotation = NULL,
    motif = NULL,
    caller = "modkit",
    min_coverage = 1L
)
```

If this works, add filters, annotation, motif scanning, and higher
`min_coverage` one at a time. That usually isolates whether the failure is in
file parsing, sample metadata, genome information, or optional annotation.

# Session Information

```{r session-info}
sessionInfo()
```
