---
title: "Working with Multiple Modification Types"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Working with Multiple Modification Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    fig.width  = 7,
    fig.height = 5,
    warning    = FALSE,
    message    = FALSE
)
```

# Introduction

Bacterial genomes frequently carry multiple simultaneous DNA methylation marks:

- **6mA** (N6-methyladenine) — deposited by Dam methyltransferase at GATC
  motifs; functions in mismatch repair and cell-cycle regulation.
- **5mC** (5-methylcytosine) — deposited by Dcm methyltransferase at CCWGG
  motifs; roles in gene regulation are less well characterized in bacteria.
- **4mC** (N4-methylcytosine) — deposited by restriction-modification systems
  at various sequence motifs; protects against foreign DNA.

`comma` stores all modification types in a single `commaData` object, enabling
joint analysis, comparison, and visualization across modification types.

This vignette demonstrates multi-modification workflows using the built-in
`comma_example_data` dataset, which contains both 6mA and 5mC sites.

```{r load}
library(comma)
data(comma_example_data)
```

# Exploring Multiple Modification Types

## What modifications are present?

```{r mod-types}
modTypes(comma_example_data)
```

## Count sites per modification type

```{r count-sites}
table(siteInfo(comma_example_data)$mod_type)
```

## Per-modification methylation summary

`methylomeSummary()` reports stats per sample. Combine it with `mod_type`
filtering to compare distributions:

```{r summary-6ma}
methylomeSummary(comma_example_data, mod_type = "6mA")[,
    c("sample_name", "condition", "mean_beta", "median_beta", "n_covered")]
```

```{r summary-5mc}
methylomeSummary(comma_example_data, mod_type = "5mC")[,
    c("sample_name", "condition", "mean_beta", "median_beta", "n_covered")]
```

# Distribution Comparison Across Modification Types

`plot_methylation_distribution()` automatically facets by modification type
when multiple types are present:

```{r plot-dist, fig.cap = "Beta value density per sample, faceted by modification type."}
plot_methylation_distribution(comma_example_data, per_sample = TRUE)
```

6mA sites in bacteria commonly show a bimodal distribution (near 0 or near 1)
because restriction-modification systems methylate recognition sequences with
high efficiency. 5mC patterns can be more variable.

# Subsetting by Modification Type

Use `subset()` to extract a single modification type:

```{r subset}
only_6ma <- subset(comma_example_data, mod_type = "6mA")
only_5mc <- subset(comma_example_data, mod_type = "5mC")

cat("6mA sites:", nrow(methylation(only_6ma)), "\n")
cat("5mC sites:", nrow(methylation(only_5mc)), "\n")
```

# Differential Methylation: Testing Each Modification Type

## Test all types in one call

By default, `diffMethyl()` tests all modification types present in the object.
Use `results(..., mod_type = "6mA")` to extract type-specific results:

```{r diffmethyl}
dm_all <- diffMethyl(comma_example_data, formula = ~ condition)
res_6ma <- results(dm_all, mod_type = "6mA")
res_5mc <- results(dm_all, mod_type = "5mC")

cat("6mA: significant sites (padj < 0.05, |Δβ| ≥ 0.2):",
    sum(!is.na(res_6ma$dm_padj) & res_6ma$dm_padj < 0.05 &
        abs(res_6ma$dm_delta_beta) >= 0.2, na.rm = TRUE), "\n")
cat("5mC: significant sites (padj < 0.05, |Δβ| ≥ 0.2):",
    sum(!is.na(res_5mc$dm_padj) & res_5mc$dm_padj < 0.05 &
        abs(res_5mc$dm_delta_beta) >= 0.2, na.rm = TRUE), "\n")
```

## Volcano plots per modification type

```{r volcanos, fig.cap = "6mA differential methylation (top) and 5mC (bottom).", fig.height = 8}
p_6ma <- plot_volcano(res_6ma) +
    ggplot2::ggtitle("6mA differential methylation")
p_5mc <- plot_volcano(res_5mc) +
    ggplot2::ggtitle("5mC differential methylation")

# Arrange side by side if patchwork is available
if (requireNamespace("patchwork", quietly = TRUE)) {
    patchwork::wrap_plots(p_6ma, p_5mc, ncol = 1L)
} else {
    print(p_6ma)
    print(p_5mc)
}
```

# PCA Across Modification Types

`plot_pca()` converts beta values to M-values via `mValues()` before PCA,
which stabilizes variance for sites near 0 or 1. By default all sites are
used; pass `mod_type` to restrict to one modification type:

```{r pca-all, fig.cap = "PCA using all modification types."}
plot_pca(comma_example_data, color_by = "condition")
```

```{r pca-6ma, fig.cap = "PCA using 6mA sites only."}
plot_pca(comma_example_data, mod_type = "6mA", color_by = "condition")
```

Use `return_data = TRUE` to retrieve the PC scores for custom plots or
downstream analysis:

```{r pca-data}
pca_df <- plot_pca(comma_example_data, mod_type = "6mA", return_data = TRUE)
# data.frame with PC1, PC2, sample_name, condition, replicate, ...
head(pca_df[, c("sample_name", "condition", "PC1", "PC2")])
attr(pca_df, "percentVar")
```

# Genome Track with Both Modification Types

When `mod_type = NULL`, `plot_genome_track()` displays sites from all
modification types, using different colors to distinguish them:

```{r genome-track, fig.cap = "Genome track showing 6mA (red) and 5mC (blue) sites."}
plot_genome_track(comma_example_data, chromosome = "chr_sim",
                  start = 1L, end = 30000L, annotation = FALSE)
```

# Heatmaps per Modification Type

Filter the `results()` output to a single modification type before passing
it to `plot_heatmap()`:

```{r heatmap-6ma, fig.cap = "Heatmap of top 20 differentially methylated 6mA sites."}
plot_heatmap(res_6ma, dm_all, n_sites = 20L)
```

# Session Information

```{r session-info}
sessionInfo()
```
