---
title: "Understanding the commaData object"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Understanding the commaData object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)
```

# Why `commaData` matters

Most commaKit workflow functions start from a `commaData` object, and many
return an updated `commaData` object with added analysis layers. Other
functions expose those layers as tables, plots, or enrichment result lists.
The object is the stable substrate for the workflow: rows are genomic
methylation sites, columns are samples, assays store measurements, and later
analyses add named layers to the same biological coordinates.

```{r load}
library(commaKit)
data(comma_example_data)

comma_example_data
```

The built-in example data are synthetic, but they exercise the same object
model used for real data. They contain multiple modification types and multiple
sequence contexts, so the examples below avoid assuming that a data set is
specific to one methylation chemistry.

# Object anatomy

`commaData` extends Bioconductor's `RangedSummarizedExperiment`. The core
dimensions therefore have the same interpretation throughout the package:

- rows are one-base methylation sites;
- columns are biological samples;
- assays are site-by-sample matrices;
- `rowRanges()` stores genomic coordinates;
- `colData()` stores sample metadata;
- `metadata()` stores package-level context such as annotations, motif sites,
  caller information, and analysis parameters.

```{r dimensions}
dim(comma_example_data)
SummarizedExperiment::assayNames(comma_example_data)
```

The main package accessors expose the two required assay matrices. Beta values
are proportions in `[0, 1]`; coverage values are read depths.

```{r assays}
methylation(comma_example_data)[1:5, 1:3]
siteCoverage(comma_example_data)[1:5, 1:3]
```

Sample metadata live in `colData()` and are also available through
`sampleInfo()`.

```{r samples}
sampleInfo(comma_example_data)
```

# Genomic row identity

Schema v2 stores genomic identity in `rowRanges()`, not in string row names.
Each row is a one-base `GRanges` range with methylation-site metadata in the
range metadata columns.

```{r row-ranges}
rr <- SummarizedExperiment::rowRanges(comma_example_data)
rr[1:4]
GenomeInfoDb::seqinfo(rr)
```

For human-readable inspection, `siteInfo()` reconstructs a flat table from
`rowRanges()` and its metadata columns.

```{r site-info}
si <- siteInfo(comma_example_data)
si[1:5, c("chrom", "position", "strand", "mod_type", "motif",
          "mod_context", "site_key")]
```

The `site_key` column is a display label. Alignment between objects and samples
is based on genomic ranges and site metadata, not on matching string keys.

# Modification type, context, and `mod_context`

The `mod_context` value should be read as **modification + context**. It is not
just a synonym for motif.

- `mod_type` answers which chemical or base modification is measured, such as
  `6mA`, `5mC`, or `4mC`.
- `motif` answers which sequence or biological context the site belongs to,
  such as `GATC`, `CCWGG`, `CpG`, or a future context definition.
- `mod_context` answers which biological class of sites is being analyzed,
  such as `6mA_GATC` or `5mC_CCWGG`.

```{r mod-contexts}
modTypes(comma_example_data)
motifs(comma_example_data)
modContexts(comma_example_data)
```

The built-in data illustrate that a `mod_context` is the combination of both
parts.

```{r context-table}
context_table <- unique(as.data.frame(si[, c("mod_type", "motif", "mod_context")]))
context_table[order(context_table$mod_context), ]
```

When a caller does not provide motif information, `motif` may be `NA` and the
computed `mod_context` falls back to the modification type alone. Code that
uses `modContexts()` rather than hard-coded labels remains compatible with
single-context, multi-context, and future-context data sets.

This distinction is important for differential methylation. Analyses should not
silently pool distinct `mod_context` groups, because methylation at different
modification-plus-context units can reflect different biological systems.

# Metadata and package-level context

Object-level context is stored in `metadata()`. In this example, annotation and
motif-site containers are available without reading any external files.

```{r metadata}
md <- S4Vectors::metadata(comma_example_data)
names(md)

c(
    annotation_features = length(annotation(comma_example_data)),
    motif_sites = length(motifSites(comma_example_data))
)

vapply(md, function(x) paste(class(x), collapse = "/"), character(1))
```

# Analysis layers

A useful way to learn commaKit is to think in layers:

- the measurement layer contains methylation and coverage assays;
- the coordinate layer is `rowRanges()`;
- the sample layer is `colData()`;
- the annotation layer adds feature relationships to sites;
- the statistical layer adds differential methylation results;
- plotting, filtering, and enrichment functions interpret those layers.

The object dimensions stay anchored to the same sites and samples while layers
are added or recomputed. The v1 policy is deliberately small and strong:

- **Raw evidence assays are canonical.** `methylation`, `coverage`,
  `mod_counts`, `canonical_counts`, and `other_mod_counts` store caller or
  reconstructed evidence. commaKit functions do not mutate these raw assays in
  place. Users can still edit assays manually with standard
  `SummarizedExperiment` tools, but package-level transformations should not
  silently rewrite the raw evidence layer.
- **Transformations are named layers.** Normalized beta values, M-values, or
  future Seurat-like transformations should be added as explicit assay layers
  with provenance and parent assays, rather than replacing the raw assays.
- **Filtering returns subset objects.** `filterSites()` and `[` follow
  Bioconductor subsetting semantics: they return a `commaData` object with
  assay layers subset to the requested rows or samples. Site-level result
  layers are trimmed for row filters; sample filters keep those site-level
  results attached to the remaining sites. commaKit does not create hidden lazy
  views.

```{r annotation-layer}
base_cols <- colnames(siteInfo(comma_example_data))

annotated <- annotateSites(comma_example_data, keep = "overlap")
annotation_cols <- setdiff(colnames(siteInfo(annotated)), base_cols)
annotation_cols

identical(dim(comma_example_data), dim(annotated))
```

Layering is intended to be explicit: rerunning an unnamed compatibility layer
updates the active/default view, while explicitly named layers can coexist for
comparison. In either case, commaKit never creates a second copy of the
biological rows.

```{r idempotent-annotation}
annotated_again <- annotateSites(annotated, keep = "overlap")
identical(colnames(siteInfo(annotated)), colnames(siteInfo(annotated_again)))
identical(dim(annotated), dim(annotated_again))
```

Differential methylation adds a statistical result layer. By default,
`diffMethyl()` works over the modification contexts present in the object,
which keeps distinct modification-plus-context units separate.

```{r diff-layer}
dm <- diffMethyl(annotated, formula = ~ condition, method = "quasi_f")
dm_cols <- setdiff(colnames(siteInfo(dm)), colnames(siteInfo(annotated)))
dm_cols

S4Vectors::metadata(dm)$diffMethyl_params[c("method", "p_adjust_method")]
resultLayers(dm)
```

`results()` exposes the active result layer as a tidy data frame while
preserving the site metadata needed to interpret each row. When you keep
multiple named runs, pass `result` or `result_name` to retrieve an older layer.

```{r results-layer}
res <- results(dm)
res <- res[order(res$dm_padj), ]
head(res[, c("chrom", "position", "strand", "mod_type", "motif",
             "mod_context", "dm_padj", "dm_delta_beta")])
```

The important pattern is that downstream functions do not replace the central
object model. They add, expose, or interpret layers attached to the same
genomic methylation-site substrate.

# Session information

```{r session-info}
sessionInfo()
```
